---
title: "List of coordinates to KML file"
author: "Hilary Rose Dawson"
date: "12/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#How to turn a list of coordinates into a KML file for Google Earth
This code is adapted from Ryan Peek's tutorial on `.gpkg` files in R. Check it out for extra spatial data information: https://ryanpeek.org/mapping-in-R-workshop/02_import_export_gpkg.html#Using_geopackage
Hattip to Christine Parisek for showing me that resource!

You can now watch these instructions as a video: https://vimeo.com/486907354

##Format your data
Create a dataframe with three columns: site name, latitude, and longitude. You can import this from a CSV or Excel file. 

sitename | latitude | longitude
---|---|---
a | 44.0 | -100.0
b | 42.0 | -101.0
c | 43.0 | -102.0

```{r}
data = read.csv(table.csv)
```

Now tell R how to read the spatial data. Use `coords` to indicate the x and y coordinates and CRS number to specify the coordinate reference system. Hint: if your CRS (or datum) is in words, search the internet for "[datum] ESPG code". 

```{r}
library(dplyr)
library(sf)
spatial = data%>%
  st_as_sf(coords=c("longitude", "latitude"), #Assign coordinate variables
           crs=4269, #Assign coordinate reference system
           remove=FALSE)
```

Check that your spatial data looks correct.
```{r}
st_crs(spatial)
```

Check that your data look right on a map.
```{r}
library(mapview)
mapview(spatial)
```

Finally, export your data to a KML.
```{r}
st_write(spatial, 
         "fia.kml", #Note the .kml ending
         layer = "FIAPlots") #Name your GIS layer
